Combined effects of protein expression variance and correlation on multicomponent systems

### 2020-03-11 Manuscript/Manuscript.Rmd
.

The coordinated expression of proteins is vital for many aspects of cellular function, from maintaining a dynamic steady state to differentiation of cells into specialized types in different tissues. The ability to regulate coordinated expression has been demonstrated to occur via transcription (transcription factors, chromatin regulation), translation (specialized ribosomes, mRNA structure/modifications), and degradation (E3 ligases). These processes, as well as the molecules they target, are all subject to noise, which has been shown to lead to differences in cell behavior and decisions in otherwise identical cells. Variability and coordinated expression (correlation) of proteins leads to increased population level control in binary decisions, and a decreased ability to execute analog signaling.

Despite these important insights, a global understanding of the variability of protein expression between single cells and the coordinated expression of groups of proteins is largely an nonexistent. To systematically assess these properties of single cells, we carried out a proteomics experiment on exceeding large single cells, Xenopus laevis eggs. An advantage of this approach is that we can get around much of the signal to noise issues that accompany studying single cells due to low sample amounts. Additionally, at this stage of development transcription is restricted so we are able to gain insights into non-transcriptional control of protein expression. We also utilized isobaric tagging in order to measure 25 individual eggs in 5 mass spectrometry runs. In this study, we were able to measure the relative abundance >1000 proteins across single cells in order to better understand the properties of stochastic and coordinated expression of proteins.

With this dataset we have been able to, for the first time, measured the relationship between protein expression variance and coordinated protein expression on a proteome scale in single cells. We have observed that certain classes of proteins, including protein complexes and metabolic pathways, are expressed in such a way that increased coordinated expression is balanced by decreased variation. By doing this, cells are able to decrease the total variance of a given complex or pathway, an elegant balancing act that allows for finer control of metabolic throughput and controlling the number of potentially formed complexes though stoichiometric control. Though this kind of coordinated expression leads to an increase in variance in a population of cells, it can reduce variation within a cell.

@@

The specific requirments for coordinated protein expression of pathways or complexes can vary, with some requiring strict stoichiometric regulation with others having much more relaxed requirements. Strict stoichiometry scenerios require coordinated expression of proteins (Figure 1A, top), and others can have uncoordinated expression (Figure 1A bottom). At the single cell level, coordinated expression will result in a high correlation coefficient, whereas un-coordinated expression will result in a low correlation coefficient (Figure 1B). Even when the expression variation (standard deviation or coefficient of variation) and abundance is identical between these two scenerios, the total variance of a pathway or complex can be significantly different due to the variance sum law: $var\left(\sum_{i=1}^n X_{i}\right) = \sum_{i=1}^{n} var(X_{i}) + 2\sum_{i,j:i<j}^{n} cov(X_{i},X_{j})$ (Figure 1C). Our previous work has shown that this effect is important in single cell pathway activation dynamics `r citep("doi.org/10.15252/msb.20177997")`.

In order to study the relationship between protein expression variance and coexpression (correlated expression) of cellular pathways and complexes at the proteome level, we utilized *Xenopus laevis* eggs as a single cell model. Activated *Xenopus laevis* eggs were collected at 5 time points across the first cell cycle (0, 20, 40, 60, and 80 minutes), with 5 eggs at each time point. Using TMT multiplexing and mass spectrometry, we were able to determine the relative abundance of more than 1300 proteins. Expression of these proteins were largely invariant across the cell cycle, revealing that these highly expressed genes are likely not regulated by cell cycle processes. Additionally, a PCA analysis of these eggs showed no discernible clustering on cell cycle time (Fig S1).

To determine the variance across these proteins, we calculated the variance and coefficient of variation (CV) using all time points collected. This showed a wide range of variation containing multiple distributions (Fig 1D), many of which are consistent with our previous study of variation using targeted mass spectrometry (Fig S2). In order to see if variation was a regulated process, we grouped the protein variation by gene ontology terms (GO terms) and plotted them in ranked order. We saw that in general, processes xxx were…

Since we were able to measure all of these proteins within single cells, we are able to calculate coordinated expression of protein pairs at single cell resolution. Using Pearson correlation coefficient, we could determine coordinated expression of nearly 2 million protein pairs (Fig 1C) The distribution of correlation coefficients fit a normal distribution, with the majority of protein pairs appearing to not show significant coregulation. However, there appear to be a significant number of protein pairs containing high correlation coefficients, and the heatmap showed a lot of clustering of these highly coregulated pairs (Fig 1D).

@@

fig_legend_1 <- c("A) Simulated expression of two proteins over time in a single cell with a mean of 100 and a standard deviation of 30 with expression correlations of 0 and 0.8. B) Pairwise plots of the two proteins in the single cells between uncorrelated and correlated scenerios. C) The total variance between the uncorrelated and correlated scenerios from A and B. D) To measure protein expression variance and coordinated expression, 25 eggs at 5 time points during the first cell cycle were analyzed with multiplexed shotgun mass spectrometry. E) Histogram of the coefficient of variation of all measured proteins. F) Histogram of the pairwise correlation coefficients between all measured proteins. G) Heatmap of the pairwise correlation coefficients between all measured protein pairs.")

@@

@@

fig_legend_2 <- c("A) Histogram of the normalized variation values across all of the GO terms and KEGG pathways. The bottom and top 10% are colored blue and red respectively. B) Pairwise plot between the mean correlation and mean coefficient of variaiton within each GO term and KEGG pathway with color coding from A. Interestingly, both the top and bottom 10% of normalized variation have groups of proteins with high levels of correlation, despite high correlation adding to total variance. In the lowest 10% of normalized variation groups there appears to be a negative correlation between mean correlation and variation.")

@@

@@

fig_legend_3 <- c("A) Histogram of the normalized variation values across all of the GO terms and KEGG pathways. The bottom and top 10% are colored blue and red respectively. B) Pairwise plot between the mean correlation and mean coefficient of variaiton within each GO term and KEGG pathway with color coding from A. Interestingly, both the top and bottom 10% of normalized variation have groups of proteins with high levels of correlation, despite high correlation adding to total variance. In the lowest 10% of normalized variation groups there appears to be a negative correlation between mean correlation and variation.")
```

## Total variance metric reveals certain modules enforce a tradeoff between protein co-expression and noise

Variance Sum Law — Dependent Case

$$var\left(\sum_{i=1}^n X_{i}\right) = \sum_{i=1}^{n} var(X_{i}) + 2\sum_{i,j:i<j}^{n} cov(X_{i},X_{j})$$

$$\sigma_{{}_{total}} = \sum_{i=1}^{n} \sigma_{X_{i}}^2 + 2\sum_{i,j:i<j}^{n} \rho(X_{i},X_{j})\sigma_{X_i}\sigma_{X_j}$$

$$\sigma_{{}_{total}} = \sqrt{var\left(\sum_{i=1}^n X_{i}\right)}$$

$$\sigma_{{}_{norm}} = \dfrac{\sigma_{{}_{total}}}{n}$$

## Enforcement of a tradeoff between protein co-expression and noise increases efficiency of metabolic pathways and heteromeric protein complexes

@@

Xenopus egg extracts were prepared based on modifications of a previous protocol (Tsai et al, 2014). All of the animal protocols used in this manuscript were approved by the Stanford University Administrative Panel on Laboratory Animal Care. To induce egg laying, female Xenopus laevis were injected with human chorionic gonadotropin injection the night before each experiment. To collect the eggs, the frogs were subjected to pelvic massage, and the eggs were collected in 1X Marc’s Modified Ringer’s (MMR) buffer (0.1 M NaCl, 2 mM KCl, 1 mM MgCl2, 2 mM CaCl2, 5 mM HEPES, pH 7.8). To remove the jelly coat from the eggs, they were placed in a solution of 2% cysteine in 1× MMR buffer for 4 min and gently agitated, after which they were washed four times with 1× MMR buffer. To activate the cell cycle, eggs were placed in a solution of 0.5 $\mu$g/ml of calcium ionophore A23187 (Sigma) and 1X MMR buffer for 3 min, after which they were washed four times with 1× MMR buffer. Single eggs were collected at their respective timepoints and placed into 600uL tubes and snap frozen in liquid nitrogen before being stored at 80°C.

@@

Single eggs were lysed mechanically by pipetting the egg in 100$\mu$L of lysis buffer (100 mM NaCl, 25 mM Tris pH 8.2, Complete EDTA- free protease inhibitor cocktail (Sigma). The lysate was then placed in a 400 uLnatural polyethylene microcentrifuge tube (E&K
Scientific #485050) and spun at 15,000 g in a right angle centrifuge (Beckman Microfuge E) at 4°C for 5 min. The lipid layer was removed by using a razor blade to cut the tube off just beneath it, and the cytoplasmic fraction was pipetted into a 1.5-ml protein LoBind tube (Fisher Scientific #13-698-794), being careful to leave the yolk behind. To precipitate the proteins from the cytoplasmic fraction, 1 ml of ice cold acetone was added to each sample and placed at 20°C overnight. To collect precipitated proteins, the samples were centrifuged at
18,000 g for 20 min at 4°C. Acetone was decanted, and the protein pellets were resolubilized in 25$\mu$L of 8 M urea. To fully solubilize the protein pellet, the samples were placed in a shaker for 1 h at room temperature. The samples were then diluted to 2 M urea with 50 mM ammonium bicarbonate to a 100$\mu$L volume, after which protein concentration was measured in duplicate with a BCA assay by taking two 10$\mu$L aliquots of each sample. The proteins in the remaining 80$\mu$L of sample volume were reduced with 10 mM TCEP and incubated for 30 min at 37°C, then alkylated with 15 mM iodoacetamide and incubated in the dark at room temperature.

@@

sup_fig_legend_1 <- c("Pairwise plots of the mean correlation coefficient and mean coefficient of variation of protien groups with linear fit. The low variance protein groups have a significant negative correlation between these two variables, showing that for protien modules that require low variance and high coexpression, showing that these two parameters... R (high, low, med) = (0.057, -0.485, 0.071)")
```

>>>>

The coordinated expression of proteins is vital for many aspects of cellular function, from maintaining a dynamic steady state to differentiation of cells into specialized types in different tissues. These requirements can range from precise stoichiometric coordination for multisubunit complexes and metabolic pathways, to high levels of noise that can result in heterogeneous responses of identical cell populations to identical stimuli `r citep("10.1073/pnas.1615660114")`. The ability to regulate coordinated expression has been demonstrated to occur via transcription (transcription factors, chromatin regulation), translation (specialized ribosomes, mRNA structure/modifications), and degradation (E3 ligases). These processes, as well as the molecules they target, are all subject to stochastic expression and interactions, which has been shown to lead to differences in cell behavior and decisions in otherwise identical cells. Variability and coordinated expression (correlation) of proteins leads to increased population level control in binary decisions, and a decreased ability to execute analog signaling.

There have been numerous targeted studies of coordinated protein expression and variation that have shown the important impacts of these parameters on cellular function. However, both of these parameters (variation and correlation) act together to determine the total variance of a system.

there is a gap in the literature of direct measurement of the relationship between variation and coordinated protein expression between single cells. To systematically assess these properties of single cells, we carried out a proteomics experiment on exceeding large single cells, Xenopus laevis eggs. An advantage of this approach is that we can get around much of the signal to noise issues that accompany studying single cells due to low sample amounts. Additionally, at this stage of development transcription is restricted so we are able to gain insights into non-transcriptional control of protein expression. We also utilized isobaric tagging in order to measure 25 individual eggs in 5 mass spectrometry runs. In this study, we were able to measure the relative abundance >1000 proteins across single cells in order to better understand the properties of stochastic and coordinated expression of proteins.

With this data set we have been able to, for the first time, measured the relationship between protein expression variance and coordinated protein expression on a proteome scale in single cells. We have observed that certain classes of proteins, including protein complexes and metabolic pathways, are expressed in such a way that increased coordinated expression is balanced by decreased variation. By doing this, cells are able to decrease the total variance of a given complex or pathway, an elegant balancing act that allows for finer control of metabolic throughput and controlling the number of potentially formed complexes though stoichiometric control. Though this kind of coordinated expression leads to an increase in variance in a population of cells, it can reduce variation within a cell.

@@

The specific requirements for coordinated protein expression of pathways or complexes can vary, with some requiring strict stoichiometric regulation with others having much more relaxed requirements. Strict stoichiometry scenarios require coordinated expression of proteins (Figure 1A, top), and others can have uncoordinated expression (Figure 1A bottom). At the single cell level, coordinated expression will result in a high correlation coefficient, whereas un-coordinated expression will result in a low correlation coefficient (Figure 1B). Even when the expression variation (standard deviation or coefficient of variation) and abundance is identical between these two scenarios, the total variance of a pathway or complex can be significantly different due to the variance sum law: $var\left(\sum_{i=1}^n X_{i}\right) = \sum_{i=1}^{n} var(X_{i}) + 2\sum_{i,j:i<j}^{n} cov(X_{i},X_{j})$ (Figure 1C). Our previous work has shown that this effect is important in single cell pathway activation dynamics `r citep("doi.org/10.15252/msb.20177997")`.

In order to study the relationship between protein expression variance and co-expression (correlated expression) of cellular pathways and complexes at the proteome level, we utilized *Xenopus laevis* eggs as a single cell model. Activated *Xenopus laevis* eggs were collected at 5 time points across the first cell cycle (0, 20, 40, 60, and 80 minutes), with 5 eggs at each time point (Fig 1D). Using TMT multiplexing and mass spectrometry, we were able to determine the relative abundance of more than 1300 proteins. Expression of these proteins were largely invariant across the cell cycle, revealing that these highly expressed genes are likely not regulated by cell cycle processes. Additionally, a PCA analysis of these eggs showed no discernible clustering on cell cycle time (Fig S2).

To determine the variance across these proteins, we calculated the variance and coefficient of variation (CV) using all time points collected. This showed a wide range of variation containing multiple distributions (Fig 1E), many of which are consistent with our previous study of variation using targeted mass spectrometry (Fig S2).

Since we were able to measure all of these proteins within single cells, we are able to calculate coordinated expression of protein pairs at single cell resolution. Using Pearson correlation coefficient, we could determine coordinated expression of nearly 2 million protein pairs (Fig 1F-G) The distribution of correlation coefficients fit a normal distribution, with the majority of protein pairs appearing to not show significant co-regulation. However, there appear to be a significant number of protein pairs containing high correlation coefficients, and the heat-map showed a lot of clustering of these highly co-regulated pairs (Fig 1G).

@@

fig_legend_1 <- c("A) Simulated expression of two proteins over time in a single cell with a mean of 100 and a standard deviation of 30 with expression correlations of 0 and 0.8. B) Pairwise plots of the two proteins in the single cells between uncorrelated and correlated scenarios. C) The total variance between the uncorrelated and correlated scenarios from A and B. D) To measure protein expression variance and coordinated expression, 25 eggs at 5 time points during the first cell cycle were analyzed with multiplexed shotgun mass spectrometry. E) Histogram of the coefficient of variation of all measured proteins. F) Histogram of the pairwise correlation coefficients between all measured proteins. G) Heat-map of the pairwise correlation coefficients between all measured protein pairs.")

@@

In order to see if variation was a regulated process, we grouped the protein variation by gene ontology terms (GO terms) and plotted them in ranked order. We saw that in general, processes xxx were…

@@

fig_legend_2 <- c("A) Histogram of the normalized variation values across all of the GO terms and KEGG pathways. The bottom and top ~15% are colored blue and red respectively. B) Distributions of the mean correlation coefficient across all groups of proteins. Both the top and bottom 15% of normalized variation groups had peaks of high levels of mean correlation coefficients. C) Distributions of the mean coefficient of variation across all groups of proteins. The top and bottom ~15% of normalized variation groups were largely separated. D) Pairwise plot between the mean correlation and mean coefficient of variation within each GO term and KEGG pathway with color coding from A. Interestingly, both the top and bottom 10% of normalized variation have groups of proteins with high levels of correlation, despite high correlation adding to total variance. In the lowest 10% of normalized variation groups there appears to be a negative correlation between mean correlation and variation.")

@@

## Total variance metric reveals certain modules enforce a trade-off between protein co-expression and noise

$$\sigma_{{}_{total}}^2 = \sum_{i=1}^{n} \sigma_{X_{i}}^2 + 2\sum_{i,j:i<j}^{n} \rho(X_{i},X_{j})\sigma_{X_i}\sigma_{X_j}$$

$$\sigma_{{}_{norm}} = \dfrac{\sqrt{\sigma_{{}_{total}}}}{n}$$

## Enforcement of a trade-off between protein co-expression and noise increases efficiency of metabolic pathways and heteromeric protein complexes

@@

fig_legend_3 <- c("A) A representative heat-map of one of the low normalized variation groups with high levels of correlated proteins (Large ribosomal subunit). B) A simple model showing the percent of potentially assembled subunits of an idealized heteromeric 10 subunit complex. Both the variation of expression between the subunits as well as the coordinated expression between them can effect the percent of maximum assembled complexes. C) Heat-map showing the total variance as a function of coefficient of variation and correlation from the idealized model in B).")
```

@@

Xenopus egg extracts were prepared based on modifications of a previous protocol `r citep("10.1371/journal.pbio.1001788")`. All of the animal protocols used in this manuscript were approved by the Stanford University Administrative Panel on Laboratory Animal Care. To induce egg laying, female Xenopus laevis were injected with human chorionic gonadotropin injection the night before each experiment. To collect the eggs, the frogs were subjected to pelvic massage, and the eggs were collected in 1X Marc’s Modified Ringer’s (MMR) buffer (0.1 M NaCl, 2 mM KCl, 1 mM MgCl2, 2 mM CaCl2, 5 mM HEPES, pH 7.8). To remove the jelly coat from the eggs, they were placed in a solution of 2% cysteine in 1× MMR buffer for 4 min and gently agitated, after which they were washed four times with 1× MMR buffer. To activate the cell cycle, eggs were placed in a solution of 0.5 $\mu$g/ml of calcium ionophore A23187 (Sigma) and 1X MMR buffer for 3 min, after which they were washed four times with 1× MMR buffer. Single eggs were collected at their respective time-points and placed into 600uL tubes and snap frozen in liquid nitrogen before being stored at 80°C.

@@

Single eggs were lysed mechanically by pipetting the egg in 100$\mu$L of lysis buffer (100 mM NaCl, 25 mM Tris pH 8.2, Complete EDTA- free protease inhibitor cocktail (Sigma). The lysate was then placed in a 400 uLnatural polyethylene micro-centrifuge tube (E&K
Scientific #485050) and spun at 15,000 g in a right angle centrifuge (Beckman Microfuge E) at 4°C for 5 min. The lipid layer was removed by using a razor blade to cut the tube off just beneath it, and the cytoplasmic fraction was pipetted into a 1.5-ml protein LoBind tube (Fisher Scientific #13-698-794), being careful to leave the yolk behind. To precipitate the proteins from the cytoplasmic fraction, 1 ml of ice cold acetone was added to each sample and placed at 20°C overnight. To collect precipitated proteins, the samples were centrifuged at
18,000 g for 20 min at 4°C. Acetone was decanted, and the protein pellets were resolubilized in 25$\mu$L of 8 M urea. To fully solubilize the protein pellet, the samples were placed in a shaker for 1 h at room temperature. The samples were then diluted to 2 M urea with 50 mM ammonium bicarbonate to a 100$\mu$L volume, after which protein concentration was measured in duplicate with a BCA assay by taking two 10$\mu$L aliquots of each sample. The proteins in the remaining 80$\mu$L of sample volume were reduced with 10 mM TCEP and incubated for 30 min at 37°C, then alkylated with 15 mM iodoacetamide and incubated in the dark at room temperature.

@@

sup_fig_legend_1 <- c("Pairwise plots of the mean correlation coefficient and mean coefficient of variation of protien groups with linear fit. The low variance protein groups have a significant negative correlation between these two variables, showing that for protein modules that require low variance and high co-expression, showing that these two parameters... R (high, low, med) = (0.057, -0.485, 0.071)")
```
